Leadership across cultures and contexts: an exploratory analysis

### 2020-06-22 leadership_across_cultures_contexts.Rmd
Editing output file.

Our data comprised one row for each of our text records. For each of our `r length(all_study_vars)` leadership dimensions, each row was coded as 1 if there was evidence in the text record for that dimension, and 0 otherwise. Because absence of evidence is not evidence of absence, our data only inform the extent to which there is evidence for each of the operationalized leadership dimensions.

Our first goal was to assess if discovery of evidence for each leadership dimensions was biased by date of publication, total pages of ethnography for a culture, and sex of the ethnographer(s). Our second goal was to assess the proportion of text records and cultures that provided evidence for each of our coded dimensions. We computed the proportion of text records supporting each dimension, and their 95% confidence intervals (CIs), using intercept-only logistic mixed effects models with random intercepts for document authors nested within cultures, fit with the lme4 package [@bates2015]. We then computed the proportion of cultures with at least one text record with evidence for each dimension, with 95% CIs computed using a cluster bootstrap.

Our third goal was to assess the extent to which each of the dimensions of leadership varied by four theoretically important factors for which we had data for all rows: group context, subsistence strategy, continental region, and leader sex. We again fit logistic mixed effects models using the lme4 package, with each leadership dimension as an outcome, all four factors entered as predictors, and author nested within culture as random intercepts. To investigate which words were associated with particular leadership dimensions, or which dimensions predicted evidence for other dimensions, we used elasticnet regression from the glmnet package [@friedman_regularization_2010]. Elasticnet models are penalized regression models that are useful when the number of predictors is large relative to the number of observations. We used the "lasso" penalty ($\alpha=1$), which will often set many coefficients to 0, thereby selecting the most important predictors.

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, minimal spanning trees and k-nearest neighbors (mst-kNN), and hierarchical cluster analysis. See the SI for descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of influential evolutionary theories of leadership. Drawing on the empirically derived features as a guide, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` documents describing `r length(unique(all_data$d_culture))` cultures.

@@

Subsistence strategy, group structure and continental region were confounded. Hunter-gatherer text records often referred to residential and kin leadership, but rarely to supracommunity leadership, whereas agriculturalist text records rarely referred to residential leadership but often referred to supracommunity leadership. Community level political groups were frequent across all subsistence types. African text records were predominantly from horticultural societies whereas North American text records were predominantly from hunter-gatherer societies. See Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence), and Table S\@ref(tab:fourfactorsumm).

@@

The majority of text records pertained to male leadership (`r male_leader_pct`%) and very few pertained to female leadership (`r female_leader_pct`%), with the remaining records not sex-specific (Table S\@ref(tab:fourfactorsumm)). There were female leaders in almost all contexts, however, except military and religious. Female leaders did tend to lead within kin and residential subgroups, though: `r female_residential_pct`% of female leaders led at the residential subgroup level, for instance, compared to `r male_residential_pct`% of male leaders. The majority of text records which exclusively described female leaders were from horticulturalists (`r female_hort_pct`%) (Table S\@ref(tab:sexsubtable)).

@@

We investigated if there were biases in the discovery of evidence for any of our `r length(all_study_vars)` leader dimensions by document publication date, female gender of any author or co-author, or total pages of ethnography from each culture. Because we found minimal evidence of bias, we do not control for any bias variables in subsequent analyses. For details, see the *Bias* section of the SI.

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (Tables S\@ref(tab:qualfuncvars) and S\@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support and then computed the proportion of cultures with at least one text record providing support for that dimension (see Methods section). See Figure \@ref(fig:varsupport) and Table S\@ref(tab:glmmcoefs).

@@

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identified as a management functions cluster (e.g., *Organize cooperation*, *Military command*, *Resolve conflict*) and the smaller of which we identified as a prosociality cluster (e.g., *Social functions*, *Distribute resources*, *Hospitality*). Within the management cluster, we identified two moderately supported subclusters: a *Strategize* subcluster (e.g., *Strategic planning*, *Provide counsel*), and an *Organize* subcluster (e.g., *Organize cooperation*, *Military command*).

We identify these seven named subclusters as "features" of leader qualities and functions in the ethnographic record. In naming these features, we gave extra weight to dimensions that were strongly supported in the sample (Figure \@ref(fig:varsupport)), and therefore not all dimensions within these clusters were clearly related to the themes of each feature. These features correspond fairly well to major elements of evolutionary theories of leadership reviewed in the introduction, with the possible exception of *Competencies*, a single feature that included at least two important theoretical elements: dominance and prestige. Because features were aggregated dimensions, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

@@

There were modest or no correlation between features, after aggregating leadership dimensions (see Figure S\@ref(fig:featurecor)). We fit separate logistic mixed effects models of each of the `r nrow(feature_models)` features as a function of group context, subsistence strategy, continental region, and leader sex, reported in the SI.

@@

Shamans and other leaders with supernatural qualities were surprisingly common in the ethnographic record. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural leader qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of *Shamanism* as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, which are a form of knowledge or expertise, and *Experienced/accomplished*). See Figure S\@ref(fig:shamanism).

>>>>

Our data comprised one row for each text record. For all `r length(all_study_vars)` leadership dimensions, each row was coded as 1 if there was evidence in the text record for that dimension, and 0 otherwise. Because absence of evidence is not evidence of absence, our data only inform the extent to which there is evidence for the operationalized leadership dimensions.

Our first goal was to assess if discovery of evidence for each leadership dimension was biased by publication date, total pages of ethnography for a culture, and sex of the ethnographer(s). Our second goal was to assess the proportion of text records and cultures that provided evidence for each of coded dimensions. We computed the proportion of text records supporting each dimension, and their 95% confidence intervals (CIs), using intercept-only logistic mixed effects models with random intercepts for document authors nested within cultures, fit with the lme4 package [@bates2015]. We then computed the proportion of cultures with at least one text record with evidence for each dimension, with 95% CIs computed using a cluster bootstrap.

Our third goal was to assess the extent to which each leadership dimensions varied by four theoretically important factors for which we had data for all rows: group context, subsistence strategy, continental region, and leader sex. We again fit logistic mixed effects models using the lme4 package, with each leadership dimension as an outcome, all four factors entered as predictors, and author nested within culture as random intercepts. To investigate which words were associated with particular leadership dimensions, or which dimensions predicted evidence for other dimensions, we used elasticnet regression from the glmnet package [@friedman_regularization_2010]. Elasticnet models are penalized regression models that are useful when the number of predictors is large relative to the number of observations. We used the "lasso" penalty ($\alpha=1$), which will often set many coefficients to 0, thereby selecting the most important predictors.

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, minimal spanning trees and k-nearest neighbors (mst-kNN), and hierarchical cluster analysis. See the SI for descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of evolutionary theories of leadership. Guided by empirically derived features, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

The geographic distribution of the cultures sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` documents describing `r length(unique(all_data$d_culture))` cultures.

@@

Subsistence strategy, group structure and continental region were confounded. Hunter-gatherer text records often referred to residential and kin leadership, but rarely to supracommunity leadership; agriculturalist text records rarely referred to residential leadership but often referred to supracommunity leadership. Community-level political groups were frequent across all subsistence types. African text records were predominantly from horticulturalits whereas North American text records were predominantly from hunter-gatherers See Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence), and Table S\@ref(tab:fourfactorsumm).

@@

The majority of text records discussed male leadership (`r male_leader_pct`%) and very few discussed female leadership (`r female_leader_pct`%), with the remaining records not sex-specific (Table S\@ref(tab:fourfactorsumm)). There were female leaders in almost all contexts, however, except military and religious. Female leaders did tend to lead within kin and residential subgroups, though: `r female_residential_pct`% of female leaders led at the residential subgroup level, compared to `r male_residential_pct`% of male leaders. The majority of text records which exclusively described female leaders were from horticulturalists (`r female_hort_pct`%) (Table S\@ref(tab:sexsubtable)).

@@

We investigated if there were biases in the discovery of evidence for any of our `r length(all_study_vars)` leader dimensions by document publication date, female gender of any author or co-author, or total pages of ethnography from each culture. We found minimal evidence of bias, and therefore do not control for bias variables in subsequent analyses. For details, see the *Bias* section of the SI.

@@

For all `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (Tables S\@ref(tab:qualfuncvars) and S\@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support and then computed the proportion of cultures with at least one text record providing support for that dimension (see Methods section). See Figure \@ref(fig:varsupport) and Table S\@ref(tab:glmmcoefs).

@@

The leader functions cluster analysis (Figure \@ref(fig:cluster)B) also identified two top-level clusters; the larger of which we identified as a management functions cluster and the smaller of which we identified as a *Prosociality* cluster. Within the management cluster, we identified two moderately supported subclusters: a *Strategize* and an *Organize* subcluster.

We identify these seven named subclusters as "features" of leadership dimensions in the ethnographic record. In naming these features, we gave extra weight to dimensions that were strongly supported empirically (Figure \@ref(fig:varsupport)), therefore not all dimensions within these clusters were clearly related to the themes of each feature. These features correspond fairly well to major elements of evolutionary theories of leadership reviewed in the introduction, with the possible exception of *Competencies*, a single feature that included at least two important theoretical elements: dominance and prestige. Because features were aggregated dimensions, they are more common at both the text record and culture level (see Figure \@ref(fig:featuressupportplot)).

@@

There were modest or no correlation between aggregageted features (see Figure S\@ref(fig:featurecor)). We fit separate logistic mixed effects models of each of the `r nrow(feature_models)` features as a function of group context, subsistence strategy, continental region, and leader sex, reported in the SI.

@@

Shamans and other leaders with supernatural qualities were surprisingly common. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural leader qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of *Shamanism* as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, a form of knowledge or expertise, and *Experienced/accomplished*). See Figure S\@ref(fig:shamanism).
